This project presents a real-time instance segmentation system for automated classification of medical tubes by cap type, developed as the upstream vision component of a wider AI-enabled medical tube segregation pipeline. An Intel RealSense D415 depth camera was used to capture a dataset of 3\,000 paired RGB and depth images across four tube classes---Universal, Push-on, Screwcap, and Other---which were annotated using SAM\,3-assisted polygon labelling on the Roboflow platform. Five segmentation models were trained and compared across three input modalities: RGB (YOLOv8m-seg, YOLO11n-seg, YOLO26n-seg), four-channel RGBD (YOLO11n-RGBD), and single-channel depth-only (YOLO26n-depth). On a held-out test split of 450 images, YOLOv8m-seg achieved the highest RGB mask mean average precision (mAP\textsubscript{50--95}) of 0.905, whilst YOLO11n-RGBD reached 0.929 by fusing depth with RGB. The depth-only model achieved 0.804 mask mAP\textsubscript{50--95}, showing that tube geometry alone provides enough information for classification---particularly useful given the reflective and transparent properties of medical tubes that challenge RGB-only approaches. YOLO26n-seg delivered the best speed--accuracy trade-off at 23.1\,ms per frame. A real-time inference pipeline was implemented, rendering segmentation masks over live RGB and depth streams at interactive frame rates. The system shows that high-accuracy tube classification is feasible with compact models suitable for edge deployment, supporting downstream conveyor-belt sorting via pneumatic actuation.
