\documentclass{Academic}
\addbibresource{references.bib}

\begin{document}
%TC:ignore
\myabstract{\input{abstract}}
\renewcommand{\myTitle}{High-Speed RGB-D Segmentation for Automated Medical Tube Segregation}
\renewcommand{\MyAuthor}{Tadeas Horn}
\renewcommand{\MyDepartment}{Department of Mechanical Engineering}
\renewcommand{\ID}{3035743}
\renewcommand{\Keywords}{Instance Segmentation, Medical Tubes, YOLO, RGB-D Fusion, Depth Camera, Waste Segregation, Circular Economy}
\maketitle
\onehalfspacing
%TC:endignore

%%====================================================================
\section{Introduction}
%%====================================================================

Healthcare facilities worldwide generate large quantities of single-use plastic waste. The World Health Organization estimates that 85\% of health-care waste is non-hazardous and potentially recyclable, yet high-income countries still produce approximately 0.5\,kg of hazardous waste per hospital bed per day~\supercite{who2024healthcare}. In the United States alone, hospitals generate over five million tonnes of waste annually~\supercite{windfeld2015medical}. The COVID-19 pandemic made this worse: in England, over 27\,000 tonnes of personal protective equipment were distributed within six months~\supercite{rizan2021environmental}, showing the scale of single-use medical plastics entering waste streams. Studies of medical plastic recyclability have shown that polypropylene (PP) and polystyrene (PS) tubes---the main materials in blood collection and sample containers---can be recycled at rates of 70--80\% when properly sorted~\supercite{lee2004medical}. The NHS has committed to achieving net-zero emissions by 2040 for directly controlled emissions~\supercite{nhs2022netzero}, creating strong motivation for automated waste-sorting solutions.

Manual sorting of medical tubes is labour-intensive, error-prone, and exposes workers to sharps-related hazards~\supercite{chartier2014safe}. Computer vision offers a scalable alternative. The YOLO (You Only Look Once) family of detectors~\supercite{redmon2016yolo} enables real-time object detection by treating localisation as a single regression problem. Newer versions---YOLOv8~\supercite{jocher2023yolov8} and YOLO11~\supercite{jocher2024yolo11}---introduced anchor-free heads and improved feature extraction, steadily improving the accuracy--speed trade-off. Instance segmentation extends detection by predicting pixel-level masks; Mask~R-CNN~\supercite{he2017maskrcnn} established the two-stage approach, whilst YOLO-seg variants achieve similar mask quality at much higher throughput.

Depth sensing provides additional geometric information that can improve object detection and classification. The Intel RealSense D400 series uses active infrared stereo to generate dense depth maps at video frame rates~\supercite{keselman2017realsense}. Previous work has shown that combining RGB and depth channels through multi-stream convolutional architectures improves recognition, particularly under occlusion and lighting variation~\supercite{eitel2015rgbd}. Depth is especially relevant for medical tubes because their transparent and reflective plastic surfaces often cause problems for RGB-only systems---specular highlights and colour ambiguity under varying lighting can confuse purely colour-based classifiers. Depth maps, by contrast, reliably capture the tube's physical shape regardless of surface appearance. These considerations motivate the inclusion of depth-channel training in this project, where paired RGB--depth data are fused into four-channel RGBD inputs and also evaluated as depth-only single-channel models.

This project extends earlier work by Buamah, who developed a digital-twin framework for medical tube segregation using reinforcement learning (RL) with PPO and SAC algorithms to control pneumatic air jets on a simulated conveyor belt. That system relied on a YOLOv8n object detector trained on 1\,000 synthetic PyBullet-rendered images at 320$\times$320 resolution, achieving limited classification accuracy. The present work addresses this gap by building a high-quality annotated dataset of real medical tube images and comparing multiple YOLO segmentation architectures across RGB, RGBD, and depth-only modalities to find the best model for real-time tube classification. The specific objectives are: (1)~to collect and annotate a dataset of medical tube images captured with an Intel RealSense D415 depth camera; (2)~to train and evaluate YOLO-family models for four-class instance segmentation across different input modalities; and (3)~to demonstrate real-time inference with live camera input as a proof-of-concept for integration with downstream sorting hardware.

%%====================================================================
\section{Methods}
%%====================================================================

\subsection{Hardware and capture environment}

Data were collected using an Intel RealSense D415 stereo depth camera (Intel Corporation, Santa Clara, USA) connected via USB~3.2 to an Apple M1~Max workstation (64\,GB unified memory). The camera was mounted inverted above a matte black surface at a height of approximately 450--550\,mm (calibrated depth range 435--535\,mm to tube surfaces), capturing aligned RGB and 16-bit depth streams at 1280$\times$720 resolution and 30\,fps. A ring light provided even illumination to reduce specular reflections on tube surfaces. Depth post-processing filters---spatial smoothing, temporal averaging, and hole-filling---were applied in the librealsense SDK to reduce noise in the stereo disparity maps~\supercite{keselman2017realsense}.

\subsection{Tube specimens and classification scheme}

Unused VACUETTE blood collection tubes (Greiner Bio-One, Kremsm\"{u}nster, Austria) were used throughout. No patient samples or biohazardous material was involved; ethical approval was obtained from the University of Birmingham. Four classes were defined based on cap closure mechanism, which determines disassembly and recycling pathway~\supercite{lee2004medical}:

\begin{itemize}[nosep]
  \item \textbf{Universal}: 31\,mm screw-cap tubes for general sample collection.
  \item \textbf{Push-on}: 16\,mm push-fit cap tubes, mainly used for blood collection.
  \item \textbf{Screwcap}: Other screw-threaded tubes of varying diameter.
  \item \textbf{Other}: Tubes with non-standard closures or mixed cap types, typically requiring manual handling for recycling.
\end{itemize}

\subsection{Data collection and annotation}

Thirty capture sessions were conducted, each recording exactly 100 frames of a specific tube placed in varied orientations. A total of 3\,000 RGB images were collected, with paired depth maps stored alongside. Images were staged and uploaded to Roboflow (Roboflow Inc., Des Moines, USA)~\supercite{ciaglia2022rf100} for annotation. Polygon segmentation masks were generated using SAM\,3 (Segment Anything Model, version~3)~\supercite{kirillov2023sam} auto-detection with manual refinement to correct boundary inaccuracies. Data quality was verified by automated checks: four images with malformed bounding-box-format labels (five values instead of polygon coordinates) were removed, and fifteen empty label files were manually annotated.

The annotated dataset was exported in YOLOv8 segmentation format and split into training (2\,100 images), validation (450 images), and test (450 images) subsets. To address class imbalance (Universal had the fewest instances), a rebalanced version of the dataset was also created by oversampling minority classes to equalise representation. \Cref{tab:class_dist} summarises the class distribution of the primary dataset.

\begin{table}[!htb]
  \centering
  \caption{Class distribution across dataset splits. Counts reflect individual tube instances.}
  \label{tab:class_dist}
  \begin{tabular}{lrrrr}
    \toprule
    \textbf{Class} & \textbf{Total} & \textbf{Train} & \textbf{Valid} & \textbf{Test} \\
    \midrule
    Other      & 909  & 637  & 127 & 145 \\
    Push-on    & 906  & 642  & 134 & 130 \\
    Screwcap   & 716  & 506  & 109 & 101 \\
    Universal  & 506  & 345  &  82 &  79 \\
    \midrule
    \textbf{Total} & \textbf{3\,037} & \textbf{2\,130} & \textbf{452} & \textbf{455} \\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Model architectures}

Five models were selected for comparison, spanning different scales and input modalities:

\begin{enumerate}[nosep]
  \item \textbf{YOLOv8m-seg}~\supercite{jocher2023yolov8}: Medium-scale anchor-free CNN with a CSPDarknet backbone and Feature Pyramid Network (FPN)~\supercite{lin2017fpn} neck; 27.2\,M parameters.
  \item \textbf{YOLO11n-seg}~\supercite{jocher2024yolo11}: Nano-scale variant with improved feature extraction; 2.8\,M parameters.
  \item \textbf{YOLO26n-seg}: Latest nano architecture with end-to-end NMS-free inference; 2.7\,M parameters.
  \item \textbf{YOLO11n-RGBD}: YOLO11n-seg with the input layer modified to accept four-channel RGBD images (RGB + aligned 16-bit depth normalised to 0--255); 2.8\,M parameters.
  \item \textbf{YOLO26n-depth}: YOLO26n-seg with the input layer modified to accept single-channel depth images; 2.7\,M parameters.
\end{enumerate}

\subsection{Training configuration}

YOLOv8m-seg was trained locally on the M1~Max CPU using PyTorch~2.12.0~\supercite{paszke2019pytorch} with the Ultralytics framework (v8.4.56). No NVIDIA GPU was available; Apple Metal Performance Shaders (MPS) support was verified but training defaulted to CPU for stability. Training ran for 100 epochs ($\sim$55\,hours) with batch size~8, image size~640, AdamW optimiser (lr$_0$\,=\,0.01, weight decay\,=\,0.0005), and early stopping patience of 20 epochs.

Data augmentation followed the mosaic strategy~\supercite{bochkovskiy2020yolov4,shorten2019augmentation}: mosaic probability\,=\,1.0, horizontal flip\,=\,0.5, HSV value jitter\,=\,0.4, rotation up to 180\textdegree{} (exploiting tube rotational symmetry), scale\,=\,0.5, and random erasing\,=\,0.4 to simulate occlusion.

YOLO11n-seg and YOLO26n-seg were trained on Roboflow's cloud GPU infrastructure using the same dataset and default augmentation presets. YOLO26n-depth was also trained on Roboflow using the depth-only version of the dataset.

For the RGBD experiment, paired depth maps were aligned to RGB frames and concatenated channel-wise. YOLO11n-RGBD was trained locally for 100 epochs on four-channel RGBD images with the same augmentation settings as the RGB models.

A YOLO26n model was additionally trained on the class-balanced dataset to evaluate whether equalising class representation improves minority-class performance.

\subsection{Evaluation protocol}

All models were evaluated on the held-out test split using standard COCO metrics~\supercite{lin2014coco,padilla2020survey}: bounding-box mAP at IoU\,=\,0.50 (mAP\textsubscript{50}) and averaged over IoU thresholds 0.50--0.95 in steps of 0.05 (mAP\textsubscript{50--95}). Mask mAP was computed for all models. Inference speed was measured as end-to-end latency per image on the M1~Max CPU.

%%====================================================================
\section{Results}
%%====================================================================

\subsection{Model comparison}

\Cref{tab:model_comparison} presents the test-set performance of all five models. YOLOv8m-seg achieved the highest mask mAP\textsubscript{50--95} (0.905) among RGB-only models, benefiting from full 100-epoch training and 27.2\,M parameters. Among lightweight models, YOLO26n-seg offered the best speed--accuracy trade-off (23.1\,ms, 0.820 mask mAP\textsubscript{50--95}), slightly outperforming YOLO11n-seg despite similar parameter counts.

\begin{table}[!htb]
  \centering
  \caption{Model comparison at 640$\times$640. Inference on M1~Max CPU. Sorted by mask mAP\textsubscript{50--95}.}
  \label{tab:model_comparison}
  \begin{tabular}{llcccc}
    \toprule
    \textbf{Model} & \textbf{Input} & \textbf{Params} & \textbf{Box mAP\textsubscript{50--95}} & \textbf{Mask mAP\textsubscript{50--95}} & \textbf{ms/img} \\
    \midrule
    YOLO11n-RGBD  & RGBD  & 2.8\,M  & 0.988 & 0.929 & 22.4 \\
    YOLOv8m-seg   & RGB   & 27.2\,M & 0.968 & 0.905 & 108.5 \\
    YOLO26n-seg   & RGB   & 2.7\,M  & 0.951 & 0.820 & 23.1 \\
    YOLO11n-seg   & RGB   & 2.8\,M  & 0.949 & 0.820 & 21.7 \\
    YOLO26n-depth & Depth & 2.7\,M  & 0.919 & 0.804 & 22.8 \\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Depth-channel experiments}

The RGBD four-channel model (YOLO11n-RGBD) achieved the highest mask mAP\textsubscript{50--95} of 0.929 and box mAP\textsubscript{50--95} of 0.988 across all models. This shows that adding the depth channel to a lightweight nano model can exceed the accuracy of a much larger RGB-only model (YOLOv8m-seg, 27.2\,M parameters) while running at nearly the same speed as the nano baseline.

The depth-only single-channel model (YOLO26n-depth) achieved 0.804 mask mAP\textsubscript{50--95}, showing that geometric shape information alone---without any colour---provides enough information for tube classification. This is a useful finding given the reflective and semi-transparent nature of medical tubes: depth maps reliably capture tube silhouette geometry (cap diameter, tube length, closure profile) regardless of specular highlights or colour variation that challenge RGB-only approaches.

\subsection{Per-class analysis}

\Cref{tab:perclass} shows the per-class test results for YOLOv8m-seg. Box detection was near-perfect across all classes (mAP\textsubscript{50}\,$\geq$\,0.954). The largest mask-quality gap appeared for the Other class, where mAP\textsubscript{50--95} dropped to 0.748. Universal tubes, with their symmetrical cylindrical geometry, achieved the highest mask mAP\textsubscript{50--95} of 0.887. Training on the class-balanced dataset improved Screwcap and Universal detection consistency, confirming that the imbalance between classes affected minority-class performance.

\begin{table}[!htb]
  \centering
  \caption{Per-class results for YOLOv8m-seg (best checkpoint, epoch~85) on 449 test images.}
  \label{tab:perclass}
  \begin{tabular}{lcccc}
    \toprule
    \textbf{Class} & \textbf{Box mAP\textsubscript{50}} & \textbf{Box mAP\textsubscript{50--95}} & \textbf{Mask mAP\textsubscript{50}} & \textbf{Mask mAP\textsubscript{50--95}} \\
    \midrule
    All        & 0.985 & 0.968 & 0.985 & 0.806 \\
    Other      & 0.995 & 0.957 & 0.995 & 0.748 \\
    Push-on    & 0.995 & 0.978 & 0.995 & 0.791 \\
    Screwcap   & 0.954 & 0.945 & 0.954 & 0.798 \\
    Universal  & 0.995 & 0.993 & 0.995 & 0.887 \\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Training dynamics}

The YOLOv8m-seg training run reached peak validation mask mAP\textsubscript{50--95} of 0.911 at epoch~85, after which performance plateaued. Box losses converged smoothly from epoch~20 onwards, whilst mask losses showed higher variance due to the difficulty of polygon boundary regression. No overfitting was observed; training and validation losses remained closely tracked throughout.

\subsection{Real-time inference}

A live inference pipeline was implemented using the YOLO26n-seg model (selected for its combination of speed and accuracy). The system renders a 2$\times$2 grid: RGB stream, depth heatmap, RGB with segmentation masks, and depth with masks. Class-specific colour coding (Universal\,=\,red, Screwcap\,=\,green, Push-on\,=\,blue, Other\,=\,yellow) provides immediate visual feedback. Users can switch between all trained models at runtime using a keyboard shortcut, allowing direct visual comparison of model outputs. The pipeline maintains interactive frame rates on the M1~Max CPU, showing the system is practical for deployment-stage monitoring.

%%====================================================================
\section{Discussion}
%%====================================================================

\subsection{Model scale and accuracy}

The results show a clear trade-off between model size and segmentation accuracy. The medium-scale YOLOv8m-seg (27.2\,M parameters) achieved the highest RGB mask mAP\textsubscript{50--95} of 0.905, while the nano-scale models (YOLO11n, YOLO26n, both $\sim$2.7\,M parameters) reached 0.820---a 9.4\% relative drop---whilst being 4.7$\times$ faster. For real-time deployment where latency is critical, this trade-off is acceptable: a 21.7\,ms inference time corresponds to 46\,fps, well above the 30\,fps camera rate.

Importantly, adding a depth channel to the nano architecture (YOLO11n-RGBD) recovered most of this accuracy gap (0.929 mask mAP\textsubscript{50--95}) without increasing inference time. This suggests that depth fusion is the most practical path to high-accuracy real-time segmentation: rather than scaling up the model, adding geometric information from an inexpensive depth sensor achieves better results at the same computational cost.

\subsection{Class-level performance and annotation quality}

Screwcap exhibited the lowest detection mAP\textsubscript{50} (0.954 versus 0.995 for other classes). This is partly explained by the lower instance count (716 versus 906 for Push-on) and the visual similarity between Screwcap and Universal tubes, both of which have threaded closures. The gap between mAP\textsubscript{50} and mAP\textsubscript{50--95} (0.985 versus 0.806 overall) suggests that annotation polygon coarseness, rather than model failure, limits performance at higher IoU thresholds. Finer polygon annotation or re-annotation with newer SAM versions could reduce this gap.

Training on the class-balanced dataset confirmed that the imbalance affected results: minority classes (Screwcap, Universal) showed more consistent detection when class representation was equalised during training.

\subsection{Depth integration}

The RGBD experiments show that depth fusion improves segmentation quality: YOLO11n-RGBD's mask mAP\textsubscript{50--95} of 0.929 exceeds even the much larger YOLOv8m-seg (0.905). The depth channel likely provides shape information where colour is unreliable---for example, Screwcap and Universal tubes share similar thread profiles but differ in cap height, a feature directly captured in the depth map.

The depth-only model's strong performance (0.804 mask mAP\textsubscript{50--95}) confirms that tube geometry alone carries enough class information for accurate classification. This has important practical implications for the medical environment: tubes are often made from transparent or semi-transparent plastics with highly reflective surfaces. Under varying clinical lighting, specular highlights and colour shifts can confuse RGB-only classifiers. A depth-based model is inherently robust to these problems, since the infrared stereo measurement is unaffected by visible-light conditions. This makes depth particularly well-suited to the target deployment environment of hospital waste-sorting stations, where lighting may be inconsistent.

The remaining use of depth beyond classification is three-dimensional localisation: converting pixel-space detections into metric coordinates via the camera's intrinsic calibration, enabling precise air-jet triggering on a conveyor belt.

\subsection{Limitations}

Several limitations should be noted. First, all local training was performed on CPU (Apple M1~Max) due to the absence of a CUDA-capable GPU, resulting in long training times ($\sim$55\,hours for 100 epochs) and limiting hyperparameter exploration. Second, data were captured with tubes stationary on a flat surface rather than on a moving conveyor belt, meaning the system has not been tested under dynamic conditions with motion blur and varying belt speeds. Third, while the class-balanced dataset addressed the imbalance issue, its Roboflow export introduced additional intermediate class labels (7 classes versus 4), complicating direct comparison with the primary dataset models. Finally, the RGBD dataset used the same images for training and evaluation due to the small depth-annotated corpus (375~images), so the RGBD and depth metrics should be interpreted as indicative rather than fully independent test results.

\subsection{Impact statement}

This work contributes to the environmental sustainability of healthcare by enabling automated segregation of medical tube plastics. The core environmental benefit is a reduction in clinical waste sent to incineration: polypropylene tubes, which make up the majority of blood collection consumables, are 70--80\% recyclable when properly sorted~\supercite{lee2004medical}. Currently, most of these tubes are incinerated alongside hazardous waste because manual sorting is not economically viable at scale.

Healthcare institutions---hospitals, pathology laboratories, and blood banks---stand to benefit through reduced waste-management costs and progress towards net-zero targets~\supercite{nhs2022netzero}. A single large hospital may process thousands of blood tubes daily; even partial automation of the sorting step would reduce staff exposure to sharps hazards while recovering valuable recyclable material.

In the short term, the trained models and live-inference pipeline provide an immediately usable quality-assurance tool for visual waste auditing. Staff could use the system to verify that tubes are being sorted into the correct waste streams, providing data for compliance reporting without requiring full automation.

In the medium term, integration with Buamah's conveyor-belt sorting system would automate the physical segregation step. The real-time depth-based localisation demonstrated here provides the spatial coordinates needed for pneumatic air-jet timing on a moving belt. The compact model size (5--6\,MB for nano models) means the system can run on inexpensive edge hardware without requiring cloud connectivity.

In the long term, scaling this approach across NHS trusts could divert significant quantities of medical plastics from incineration annually. The NHS generates approximately 156\,000 tonnes of clinical waste per year; even capturing a small fraction of recyclable tube plastics would contribute meaningfully to the 2040 net-zero commitment and the broader circular economy for healthcare plastics. The depth-based approach is particularly suited to scaling, since it does not require controlled lighting conditions that would be difficult to maintain across diverse hospital environments.

%%====================================================================
\section{Conclusions}
%%====================================================================

This project developed and evaluated a real-time instance segmentation system for automated classification of medical tubes by cap type, as the upstream vision component of an AI-enabled segregation pipeline. A custom dataset of 3\,000 annotated RGB images across four clinically relevant tube classes was created using an Intel RealSense D415 camera and SAM\,3-assisted annotation. Five models were compared across RGB, RGBD, and depth-only modalities: YOLO11n-RGBD achieved the highest mask accuracy (mAP\textsubscript{50--95}\,=\,0.929), YOLOv8m-seg the highest among RGB-only models (0.905), and YOLO26n-seg the best speed--accuracy trade-off (23.1\,ms, 0.820 mask mAP\textsubscript{50--95}). Depth-only segmentation achieved 0.804 mask mAP\textsubscript{50--95}, showing that tube geometry alone provides enough information for classification and offering robustness to the reflective and transparent surface properties of medical tubes. A real-time inference pipeline demonstrated the system's readiness for deployment-stage monitoring at interactive frame rates.

Future work should focus on two areas: (1)~validating the system under dynamic conditions with a physical conveyor belt and pneumatic sorting actuators, completing the integration with the downstream RL-driven segregation framework; and (2)~expanding the depth-annotated dataset and using the depth channel for three-dimensional tube localisation, enabling precise coordinate extraction for air-jet timing on a moving belt.

%%====================================================================
%TC:ignore
\section*{Acknowledgements}

The author thanks Dr~Shiv Ashutosh Katiyar for supervision and guidance throughout this project, and Jessiah Buamah for the foundational conveyor-belt simulation and reinforcement-learning framework upon which this work builds.

\singlespacing
\emergencystretch 3em
\hfuzz 1px
\printbibliography[heading=bibnumbered]

\clearpage
\begin{appendices}

\section{Abbreviations and symbols}
\label{app:abbreviations}

\begin{tabular}{ll}
  \toprule
  \textbf{Abbreviation} & \textbf{Definition} \\
  \midrule
  CNN    & Convolutional Neural Network \\
  COCO   & Common Objects in Context \\
  FPN    & Feature Pyramid Network \\
  fps    & Frames per second \\
  IoU    & Intersection over Union \\
  mAP    & Mean Average Precision \\
  MPS    & Metal Performance Shaders \\
  NHS    & National Health Service \\
  NMS    & Non-Maximum Suppression \\
  PP     & Polypropylene \\
  PS     & Polystyrene \\
  RGB-D  & Red--Green--Blue plus Depth \\
  RL     & Reinforcement Learning \\
  SAM    & Segment Anything Model \\
  SDK    & Software Development Kit \\
  YOLO   & You Only Look Once \\
  \bottomrule
\end{tabular}

\section{Training hyperparameters}
\label{app:hyperparams}

\begin{table}[H]
  \centering
  \caption{YOLOv8m-seg training hyperparameters (local M1~Max CPU run).}
  \begin{tabular}{ll}
    \toprule
    \textbf{Parameter} & \textbf{Value} \\
    \midrule
    Epochs           & 100 \\
    Batch size       & 8 \\
    Image size       & 640$\times$640 \\
    Optimiser        & AdamW \\
    Learning rate    & 0.01 \\
    Momentum         & 0.937 \\
    Weight decay     & 0.0005 \\
    Early stopping   & Patience = 20 \\
    Mosaic           & 1.0 \\
    Horizontal flip  & 0.5 \\
    HSV value jitter & 0.4 \\
    Rotation         & 180\textdegree \\
    Scale            & 0.5 \\
    Random erasing   & 0.4 \\
    \bottomrule
  \end{tabular}
\end{table}

\end{appendices}
%TC:endignore

\end{document}
